\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-21 paper.tex
Add intro to application data sets

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

In this article, mainly focus on five NLDR techniques. They are tSNE,
UMAP, PHATE, TriMAP \citep{article02}, and Pairwise Controlled Manifold
Approximation (PaCMAP) \citep{article01}.

@@

PaCMAP \citep{article01} is different because it adds supervised
learning to make a 2D representation while keeping the relationships
between pairs of points. It builds a graph using distances between pairs
and then makes the 2D representation better using a customizable loss
function. What's special about PaCMAP is that it can use class labels or
extra information to guide how it makes the 2D representation. This
gives users a way to change how PaCMAP works to fit their needs better.

@@

Typically, the MSE plot tends to have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope gradually declines or levesl off, indicating that
a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. Figure~\ref{fig-diagnosticpltScurve} (b) has created a
graph that shows how the Mean Squared Error (MSE) changes depending on
the number of bins assigned in the hexagonal grid for UMAP applied to
the S-curve dataset. As per the rule of thumb mentioned earlier, the
effective number of bins is determined to be \(84\), with \(6\) bins
along the x-axis and \(14\) bins along the y-axis. Furthermore,
Figure~\ref{fig-modelScurve} shows the 2D model constructed using UMAP
with the selected total number of bins and how the model overlay on
high-D data.

@@

After setting up the hexagonal grid with an appropriate number of bins,
it is possible that some hexagonal bins may have very few or no data
points within them. To ensure comprehensive coverage of the NLDR data,
it is necessary to select hexagonal bins that have a considerable number
of data points within their hexagons. To achieve this, the standard
number of points within each hexagon is first calculated. This standard
count is obtained by dividing the number of points within each hexagon
by the maximum number of points in the grid (see
Equation~\ref{eq-equationp2}). Next, the bins that have a standard
number of points less than a certain benchmark value are removed. After
removing the hexagons that have insufficient data density, the hexagons
with more substantial data representation are used to construct the 2D
model.

@@

Furthermore, it is crucial to choose the benchmark value to remove
low-density hexagons carefully. If we remove unnecessary bins, then it
may result in long edges and an uneven 2D model. Therefore, instead of
removing hexagons only identified by the benchmark value, we examine the
standard number of points in the neighboring hexagons of the identified
low-density hexagons. If the neighboring bins also have low counts, then
only those bins will be removed. The other identified bins will be kept
and used for constructing the 2D model along with the high-density bins
(see Appendix for more details).

The benchmark value for removing low-density hexagons has a range of
\(0\) and \(1\). When analyzing how these benchmark values affect model
performance, it's important to observe the change in Mean Squared Error
(MSE) as the benchmark value increases. The MSE exhibits a gradual
increase as the benchmark value progresses from \(0\) to \(1\).
Assessing this rate of increase is crucial. If the increment is not
substantial, the decision might lean towards retaining low-density
hexagons. According to \textbf{?@fig-diagnosticpltScurvelwd}, the change
in MSE for NLDR techniques is relatively small across different
benchmark values, except for the PHATE method.

Consequently, it is not necessary to remove low-density hexagons when
constructing the 2D model for the S-curve dataset in NLDR techniques,
except for PHATE. While there may be some fluctuations in MSE between
benchmark values of 0 to 1, the absence of significant changes suggests
that maintaining low-density hexagons is preferable. However, to choose
the best selection, it's recommended to examine the benchmark values
around the first local minimum in the MSE curve. However, the MSE for
PHATE fluctuates when benchmark values range between 0 and 1. Therefore,
it is necessary to explore the benchmark values surrounding the first
local minimum to determine an effective benchmark value. There is no
universal rule for selecting a process. As shown in
\textbf{?@fig-diagnosticpltScurvelwd}, the MSE values for PHATE
fluctuate, and the first local minimum is at \(0\). Therefore, nearby
benchmark values such as \(0.01\), \(0.04\), and \(0.06\) were explored
to identify the best value for preserving the data structure. In this
case, a benchmark value of \(0.06\) was chosen. The resulting 2D model
is shown in Figure~\ref{fig-modelScurve} (c), while the video link
provides a visual representation of how the model appears in high-D.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Plot the distribution of the standardized counts
\item
  See the distribution of counts
\item
  Take the first quartile
\end{enumerate}

@@

@@

\caption{\label{fig-modelScurve}(a), (b) 2D model.}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Plot the distribution of the 2D distances
\item
  Find a value which is greater than smallest value
\end{enumerate}

@@

To assess how well our method captures and represents the underlying
structure of the high-D data, Mean Squared Error (MSE) and Akaike
Information Criterion (AIC) are used. When computing MSE, total model
error (see Section~\ref{sec-summary}) is divided by the number of
observations to make it as a mean value (see
Equation~\ref{eq-equation9}).

@@

Furthermore, to incorporate with the parameters in our model, which is
the non-empty bins, and as a statistical measure used for model
selection and comparison, AIC is used (see
Equation~\ref{eq-equation10}).

\begin{equation}\protect\hypertarget{eq-equation10}{}{
\text{AIC} = 2b'p + np * log(\text{MSE})
}\label{eq-equation10}\end{equation}

Here, \(b'\) signifies the number of bins non-empty bins, \(p\) denotes
the number of dimensions in the high-D data, and \(n\) represents the
number of observations.

@@

In the field of single-cell studies, a common analysis task involves
clustering to identify groups of cells with similar expression profiles.
Analysts often turn to NLDR techniques to verify and identify these
clusters and explore developmental trajectories. In clustering
workflows, the main objective is to verify the existence of clusters and
subsequently identify them as specific cell types by examining the
expression of ``known'' marker genes. Marker genes are specific genes or
transcripts that are uniquely or highly expressed in particular cell
types. Marker genes are identified through differential gene expression
(DEG) analysis, comparing the expression levels of genes across
different cell types. In this context, a ``faithful'' embedding should
ideally preserve the topology of the data, ensuring that cells
corresponding to the same cell type are situated close to the
high-dimensional space.

\hypertarget{processing-steps}{%
\subsection{Processing steps}\label{processing-steps}}

To begin our analysis, we installed the Peripheral Blood Mononuclear
Cells (pbmc) data set obtained from 10x Genomics using the
\texttt{SeuratData} R package \citep{Rahul2019}, which facilitates the
distribution of data sets in the form of Seurat objects
\citep{Yuhan2021}. This data set contains 13,714 features (genes) across
2,700 samples (cells) within a single assay. The active assay is RNA,
with 13,714 features representing different gene expressions. Then, the
cells that have unique feature counts over 2,500 or less than 200 and
cells that have \textgreater{} 5\% mitochondrial counts are filtered.
After removing unwanted cells from the dataset, the next step is to
normalize the data. By default, we employ a global-scaling normalization
method ``LogNormalize'' that normalizes the feature expression
measurements for each cell by the total expression, multiplies this by a
scale factor (10,000 by default), and log-transforms the result.

Next, top 1000 genes are selected by Festem \citep{Chen2023} DEG method.
Then, scale the data. Then, perform PCA on the scaled data. Louvain
algorithm is used to cluster the single cells based on the genes
selected by Festem. Using the Festem-selected genes and 15 PCs,
identified 10 clusters in the PBMC3k data and annotated them based on
the expression of canonical markers. The 10 clusters included immune
cells such as naive CD4 T cells, memory CD4 T cells, CD8 T cells, CD14
monocytes, FCGR3A monocytes, Natural Killer (NK) cells, B cells and
Dendritic Cells (DC). In addition to these common cell types, Festem
identified a fine cell type, CD27− CD4+ memory T cells, which were often
missed by other methods.

\begin{itemize}
\item
  Human peripheral blood mononclear cells (PBMCs) data
\item
  Intro to the data set
\end{itemize}

Used Louvain algorithm to cluster the single cells based on the genes
selected by Festem method

\begin{itemize}
\item
  Explain what ``Evaluate Festem on DEG detection'' means, in terms of
  this data
\item
  How did they ``identified 10 clusters''? What were the variables used,
  PCs or full data or 2D representation, spell this out.
\item
  Used top 1000 genes selected by Festem to run PCA, top 15 PCs
  (original paper mentioned)
\item
  The ``10 clusters included immune cells such as naive CD4 T \ldots{}''
  does this mean that every cluster had all of these, or one cluster was
  mostly ``CD4 T cells'', another mostly ``memory CD4 T'', \ldots?
\item
  How is your use of this data different from the original application.
\item
  Original paper used this data set to compare different DEG methods,
  but we need to look how the performance of UMAP with the author's
  selected parameter choice.
\item
  Why 15 PCs? Is that from the original paper? YES
\end{itemize}

\citet{Chen2023}

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\item
  15 PCs
\item
  Num bins along the x-axis = 23, shape parameter = 0.8772751
\item
  learned from the model: There are 3 well-separated clusters in 2D, but
  if look at the model in high-D space, the three clusters are much
  closer. Also, there is some continuity within the clustering structure
  and it didn't capture well. Therefore UMAP with n:neighbour 30 is not
  a good representation for this data set.
\item
  The most important parameter is n\_neighbors - the number of
  approximate nearest neighbors used to construct the initial
  high-dimensional graph. It effectively controls how UMAP balances
  local versus global structure - low values will push UMAP to focus
  more on local structure by constraining the number of neighboring
  points considered when analyzing the data in high dimensions, while
  high values will push UMAP towards representing the big-picture
  structure while losing fine detail.
\item
  The second parameter we'll investigate is min\_dist, or the minimum
  distance between points in low-dimensional space. This parameter
  controls how tightly UMAP clumps points together, with low values
  leading to more tightly packed embeddings. Larger values of min\_dist
  will make UMAP pack points together more loosely, focusing instead on
  the preservation of the broad topological structure.
\item
  The third parameter is metric. This parameter plays a crucial role in
  determining how distances are computed within the ambient space of the
  input data (The term ``ambient space'' refers to the original or
  initial space in which the data exists before any transformation or
  dimensionality reduction).
\item
  If it is completely a mess, then that is also further evidence that
  the fit is poor in high-dimensions.
\end{itemize}

\hypertarget{approach-followed}{%
\subsection{Approach followed}\label{approach-followed}}

\begin{itemize}
\item
  First, referred to \citet{Chen2023} paper which shows UMAP
  representation of PBMC3k data with 15 PCs and parameter choice of
  n:neighbor:30, min\_dist:0.1, metric:cosine.
\item
  Observed that there is three well-separated clusters.
\item
  Second, visualise the 15 PCs in high-dimensional space with our model
  and investigate whether can I see the three well-separated clusters or
  is it showing something different than what I saw in UMAP view
  suggested by the author.
\item
  Learned from the model: Yes, it's different. There are three clusters
  but they are really close to each other. Also, in the largest cluster,
  observed non-linear continuity structure, which didn't captured by the
  author's UMAP representation.
\item
  Learned from the data: Visualise with detour and colored the clusters
  in high-D. Identified more than three clusters which is suprising.
\item
  In that case, need to suggest a better UMAP representation than
  author's suggestion.
\item
  To find a better UMAP representation, I need to investigate different
  parameter choices for UMAP (or need to try another NLDR technique, but
  here my objective is to suggest the best parameter choice with UMAP).
\item
  Therefore, I fitted our model for different UMAP parameter choices,
  record the error, and selected the parameter choice that gives the
  lowest error which is n:neighbor:5, min\_dist:0.99, metric:cosine.
\item
  Then, construct our model for this UMAP data (n:neighbor:5,
  min\_dist:0.99, metric:cosine).
\item
  In this representation, you can see that the clusters are closer, and
  the continuity of clusters are there.
\item
  Also, what missed from looking at high-D data with detour, in this
  UMAP representation I can see more than three clusters.
\item
  Then, need to check how correct am I identify the clusters in 2D and
  is that matched that I identify in high-D.
\item
  Generate a confusion matrix.
\end{itemize}

\hypertarget{story}{%
\subsection{Story}\label{story}}

\begin{itemize}
\tightlist
\item
  UMAP representation of PBMC3k data with 15 PCs and parameter choice of
  n:neighbor:30, min\_dist:0.1, metric:cosine (author's suggestion)
  shows three well-separated clusters (see
  Figure~\ref{fig-nldervisPBMCUMPview}). The question we need to answer
  is: Does this UMAP representation preserve the actual data structure,
  or is something missing or misidentified or misleading?
\end{itemize}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMPview-1.pdf}

@@

\caption{\label{fig-nldervisPBMCUMPview}(a) 2D layout from UMAP
(n\_neighbors = 30, min\_dist = 0.3) applied for the PBMC3k dataset. Is
this a best representation of the original data?}

@@

\begin{itemize}
\item
  To investigate this, we are going to visualise the model in the data
  space, which means that the model generated from UMAP is visualised in
  high-D with original data.
\item
  To fit the model, first we need to decide how many number bins should
  allocate along the x and y axes. To make an initial choice, we use MSE
  plot (Figure~\ref{fig-diagnosticpltPBMC} (b)).
\end{itemize}

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltPBMC-1.pdf}

@@

\caption{\label{fig-diagnosticpltPBMC}Goodness of fit statistics from
UMAP applied to training PBMC3k dataset. What is the effective number of
bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
588, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

@@

\begin{itemize}
\tightlist
\item
  Then, the model is constructed by following all the steps in 2D and
  lifts the model into high-D.
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  As shown in Figure~\ref{fig-nldervisPBMCUMAP}, there are three
  well-separated clusters (according to the model in 2D space).
\end{itemize}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_1.png}

@@

\subcaption{\label{fig-pbmc1_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_2.png}

\subcaption{\label{fig-pbmc1_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_3.png}

\subcaption{\label{fig-pbmc1_sc3}}
\end{minipage}%

\caption{\label{fig-pbmc1_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/Gnhh4hfsbbg}).}

@@

\begin{itemize}
\item
  According to Figure~\ref{fig-pbmc1_sc} (a), the model in high-D space
  shows two well separated clusters. Also, two clusters are really close
  to each other (see Figure~\ref{fig-pbmc1_sc} (b)). Furthermore, there
  are three clusters shown by the model in 2D actually represented in
  high-D as well (see Figure~\ref{fig-pbmc1_sc} (c)). Another important
  thing to notice from the model in high-D space is that the non-linear
  continuity structure in two well-separated clusters.
\item
  Figure~\ref{fig-pbmc1_sc} (c) gives evidence that UMAP preserves the
  global structure. Furthermore, the non-linearity within the clusters
  shown in Figure~\ref{fig-pbmc1_sc} (a) proves that local structure is
  also preserved but that's not visually shown by th UMAP layout in 2D.
\item
  Can we find a better parameter choice for UMAP which shows this
  non-linear continuity of the big clusters in 2D space as well?
\item
  To do that, we are going to fit models for different parameter choice
  and select the one which have the lowest model error in high-D (see
  \textbf{?@tbl-errordiffcomb}).
\end{itemize}

\begin{table}[H]

\caption{errors for different parameter combinations}
\centering
\begin{tabular}[t]{r|r|l|r}
\hline
\textbf{n\_neighbors} & \textbf{min\_dist} & \textbf{metric} & \textbf{error (x 100)}\\
\hline
\cellcolor{gray!6}{5} & \cellcolor{gray!6}{0.99} & \cellcolor{gray!6}{cosine} & \cellcolor{gray!6}{262.18}\\
\hline
15 & 0.99 & cosine & 279.49\\
\hline
\cellcolor{gray!6}{84} & \cellcolor{gray!6}{0.99} & \cellcolor{gray!6}{cosine} & \cellcolor{gray!6}{295.64}\\
\hline
5 & 0.30 & cosine & 303.90\\
\hline
\cellcolor{gray!6}{15} & \cellcolor{gray!6}{0.30} & \cellcolor{gray!6}{cosine} & \cellcolor{gray!6}{308.97}\\
\hline
84 & 0.30 & cosine & 319.59\\
\hline
\textcolor{red}{\textbf{\cellcolor{gray!6}{30}}} & \textcolor{red}{\textbf{\cellcolor{gray!6}{0.30}}} & \textcolor{red}{\textbf{\cellcolor{gray!6}{cosine}}} & \textcolor{red}{\textbf{\cellcolor{gray!6}{386.35}}}\\
\hline
\end{tabular}
\end{table}

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMAPview2-1.pdf}

\caption{\label{fig-nldervisPBMCUMAPview2}(a) 2D layout from UMAP
(n\_neighbors = 5, min\_dist = 0.99, metric = cosine) applied for the
PBMC3k dataset. Is this a best representation of the original data?}

\begin{itemize}
\item
  Selected the one with n:neighbor:5, min\_dist:0.99, metric:cosine (see
  Figure~\ref{fig-nldervisPBMCUMAPview2}). As shown in
  Figure~\ref{fig-nldervisPBMCUMAPview2}, the two big clusters have
  non-linear continuity structures, two big clusters are well-separated
  but close to each other, and the two cluster which exists in the one
  big cluster are really close to each other (see
  Figure~\ref{fig-nldervisPBMCUMAPview2}).
\item
  Now, we are going to fit the model for the new parameter choice.
\end{itemize}

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltPBMC3-1.pdf}

\caption{\label{fig-diagnosticpltPBMC3}Goodness of fit statistics from
UMAP applied to training S-curve dataset. What is the effective number
of bins in each NLDR technique to create a 2D model? The MSE plot have a
steep slope at the beginning, indicating that a smaller number of bins
causes a larger amount of error. Then, the slope gradually declines or
level off, indicating that a higher number of bins generates a smaller
error. Using the elbow method, when the total number of bins is set to
588, the slope of the Mean Squared Error (MSE) plot experiences a sudden
and noticeable change, resembling an elbow-like shape. This point
indicates that adding less bins does not enough to capture the data
structure.}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisPBMCUMAP3-1.pdf}

\caption{\label{fig-nldervisPBMCUMAP3}(a) Model in the 2D space overlaid
with UMAP data, and (b) Error occurred by the high-D model for high-D
data and the error mapped into 2D NLDR points.}

\begin{itemize}
\tightlist
\item
  According to Figure~\ref{fig-nldervisPBMCUMAP3} (a), the model
  constructed in 2D space for the new parameter choice of UMAP shows 5
  clusters which are really close to each other. The big cluster shows
  the non-linear continuity structure with two sub clusters. Other three
  clusters positioned in a structure of non-linearity but really close
  to each other.
\end{itemize}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_4.png}

\subcaption{\label{fig-pbmc2_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_5.png}

\subcaption{\label{fig-pbmc2_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_6.png}

\subcaption{\label{fig-pbmc2_sc3}}
\end{minipage}%

\caption{\label{fig-pbmc2_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D with new parameter choice
for UMAP, a video of the tour animation is available at
(\url{https://youtu.be/lc_6-rIuNHI}).}

\begin{itemize}
\item
  As shown in Figure~\ref{fig-pbmc2_sc}, there are 5 clusters, three of
  them are well-separated but close to each other as seen in
  Figure~\ref{fig-nldervisPBMCUMAP3} (a). Also, other two clusters are
  really close to each other but separated from the early mentioned
  three clusters.
\item
  Next, we are going to investigate whether, the clusters we identified
  in 2D UMAP layout, can also be identified in high-D as well.
\item
  First, I colored the clusters in 2D layout manually. Then, I used
  spin-and-brush approach with \texttt{detour} to color the clusters in
  high-D.
\end{itemize}

\begin{verbatim}

            cluster 1 cluster 2 cluster 3 cluster 4 cluster 5 cluster 6
  cluster 1       138        13         0         0         0         0
  cluster 2        11       509         0         1         0         0
  cluster 3         1         0       348         1         0         1
  cluster 4         0         0         3      1029        64         2
  cluster 5         0         0         0        85       226        32
  cluster 6         0         0         0         0        19       139
\end{verbatim}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisPBMCUMAP3Res-1.pdf}

\caption{\label{fig-nldervisPBMCUMAP3Res}2D layout from UMAP
(n\_neighbors = 5, min\_dist = 0.99, metric = cosine) applied for PBMC3k
dataset (a) colored manually in 2D, (b) colored in high-D.}

\begin{itemize}
\item
  Figure~\ref{fig-nldervisPBMCUMAP3Res} shows the 2D UMAP layout colored
  with each method. As shown in Figure~\ref{fig-nldervisPBMCUMAP3Res}
  (b) and the confusion matrix,there are some miss-classifications. What
  is the reason for that?
\item
  For example, the point (ID: 1158) classified as \textbf{cluster 3} in
  2D UMAP layout, classified as \textbf{cluster 6} when doing the
  colouring in high-D (see Figure~\ref{fig-nldervisPBMCUMAP3Res}). Why?
\end{itemize}

\begin{minipage}[t]{\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_7.png}

\subcaption{\label{fig-pbmc2_sc7}}
\end{minipage}%

\caption{\label{fig-pbmc2_appsc}Screen shots of app which shows 2D UMAP
layout and comparison of distances related to the point classified as
\textbf{cluster 6} in high-D (ID: 1158) and randomly selected point
which is in \textbf{cluster 6} (ID: 2291).}

\begin{itemize}
\tightlist
\item
  If we compute the distance between these two points, you will see that
  it preserves the pair-wise distance (see
  Figure~\ref{fig-pbmc2_appsc}), which give us a evidence that even
  though the point 1158, classified as \textbf{cluster 3} in 2D layout,
  actually should be in \textbf{cluster 6}. This also evident that some
  misleading can happens with UMAP.
\end{itemize}

\begin{minipage}[t]{\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_8.png}

\subcaption{\label{fig-pbmc2_sc8}}
\end{minipage}%

\caption{\label{fig-pbmc2_appsc1}Screen shots of app which shows 2D UMAP
layout and comparison of distances related to the point classified as
\textbf{cluster 5} in high-D (ID: 3) and randomly selected point which
is in \textbf{cluster 5} (ID: 2198).}

\begin{itemize}
\tightlist
\item
  For another example, if we compute the distance between these two
  points, you will see that it preserves the pair-wise distance (see
  Figure~\ref{fig-pbmc2_appsc1}), which give us a evidence that even
  though the point 3, classified as \textbf{cluster 4} in 2D layout,
  actually should be in \textbf{cluster 5}. This also evident that some
  misleading can happens with UMAP.
\end{itemize}

\begin{itemize}
\item
  Why I choose this data set? Because I need show that whatever NLDR
  technique that we use, our model can capture the non-linear structure
  within clusters as well.
\item
  Digit 1 (7877 images only)
\end{itemize}

\hypertarget{preprocessing}{%
\subsubsection{Preprocessing}\label{preprocessing}}

\begin{itemize}
\item
  Scale the pixel values to the range {[}0,1{]}
\item
  Reshape to have a 2-dimensional array
\item
  reshaping pixel data is a fundamental preprocessing step in image
  analysis tasks, ensuring that the data is in a suitable format for
  further analysis, modeling, and interpretation. It helps preserve
  important spatial information, ensures compatibility with machine
  learning models, and facilitates feature extraction and visualization.
\item
  Reshaping the pixel data allows us to preserve this spatial
  information, ensuring that the relationships between neighboring
  pixels are maintained.
\item
  Many machine learning algorithms, including neural networks, expect
  input data to be in a specific format. For image data, this typically
  means a two-dimensional array where each row represents a single
  sample (image), and each column represents a feature (pixel value).
  Reshaping ensures that all samples have consistent dimensions, making
  them suitable for feeding into these algorithms.
\item
  Reshaping the pixel data allows for easier extraction of relevant
  features from the images. Once reshaped, individual pixels can be
  treated as features, making it easier to apply techniques like
  dimensionality reduction (e.g., PCA) or feature engineering to extract
  meaningful information from the images.
\item
  Selected 10 PCs based on scree plot
\item
  7877 images of digit 1
\item
  PaCMAP parameters: n\_components=2, n\_neighbors=10, init=``random'',
  MN\_ratio=0.9, FP\_ratio=2.0
\end{itemize}

\hypertarget{model}{%
\subsubsection{Model}\label{model}}

\begin{itemize}
\item
  Total number of bins used: 180 (num\_x = 12, num\_y = 15, hex\_size =
  0.06)
\item
  Learned from the model: There is a non-linear shape in 2D, but if look
  at the model in high-D, the non-linear continuity is there but there
  is a twist in the structure which didn't capture by PaCMAP in 2D
  layout.
\end{itemize}

\hypertarget{story-1}{%
\subsubsection{Story}\label{story-1}}

\begin{itemize}
\item
  PaCMAP representation of MNIST digit 1 data with 10PCs and parameter
  choice of n\_components=2, n\_neighbors=10, init=``random'',
  MN\_ratio=0.9, FP\_ratio=2.0 shows a non-linear structure (see
  Figure~\ref{fig-nldervisMNISTPACMAP} (a)). I selected this parameter
  choice because this the default parameter setting for PaCMAP and
  selected PaCMAP because when I observed the 5 NLDR techniques applied
  for the data PaCMAP showed the non-linear structure effectively in 2D.
\item
  By looking at MSE plot, I selected the 180 bins, which associated with
  12 bins along the x-axis, 15 bins along the y-axis, and hex size as
  0.06 (Figure~\ref{fig-diagnosticpltMNIST} (b)). Then, fitted the model
  in 2D (see Figure~\ref{fig-nldervisMNISTPACMAP} (b)). As you can see
  in Figure~\ref{fig-nldervisMNISTPACMAP} (b), the fitted model capture
  the non-linear structure of the 2D data.
\end{itemize}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

@@

In this article, mainly focus on five NLDR techniques. They are tSNE,
UMAP, PHATE, TriMAP \citep{article02}, and Pairwise Controlled Manifold
Approximation (PaCMAP) \citep{Yingfan2021}.

@@

PaCMAP \citep{Yingfan2021} is different because it adds supervised
learning to make a 2D representation while keeping the relationships
between pairs of points. It builds a graph using distances between pairs
and then makes the 2D representation better using a customizable loss
function. What's special about PaCMAP is that it can use class labels or
extra information to guide how it makes the 2D representation. This
gives users a way to change how PaCMAP works to fit their needs better.

@@

Typically, the MSE plot tends to have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope gradually declines or levesl off, indicating that
a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. Figure~\ref{fig-diagnosticpltScurve} has created a graph
that shows how the Mean Squared Error (MSE) changes depending on the
number of bins assigned in the hexagonal grid for UMAP applied to the
S-curve dataset. As per the rule of thumb mentioned earlier, the
effective number of bins is determined to be \(84\), with \(6\) bins
along the x-axis and \(14\) bins along the y-axis. Furthermore,
Figure~\ref{fig-modelScurve} shows the 2D model constructed using UMAP
with the selected total number of bins and how the model overlay on
high-D data.

@@

After setting up the hexagonal grid with an appropriate number of bins,
it is possible that some hexagonal bins may have very few or no data
points within them. To ensure comprehensive coverage of the NLDR data,
it is necessary to select hexagonal bins that have a considerable number
of data points within their hexagons. To achieve this, the number of
points within each hexagon is first calculated. Then, the standard count
is computed by dividing the number of points within each hexagon by the
maximum number of points in the grid (see Equation~\ref{eq-equationp2}).
Next, the bins that have a standard number of points less than a certain
benchmark value are removed. After removing the hexagons that have
insufficient data density, the hexagons with more substantial data
representation are used to construct the 2D model.

@@

Furthermore, it is crucial to choose the benchmark value to remove
low-density hexagons carefully. If we remove unnecessary bins, then it
may result in long edges and an uneven 2D model. Therefore, instead of
removing hexagons only identified by the benchmark value, we examine the
standard number of points in the neighboring hexagons of the identified
low-density hexagons. If the neighboring bins also have low counts, then
only those bins will be removed. The other identified bins will be kept
and used for constructing the 2D model along with the high-density bins.

The benchmark value for removing low-density hexagons has a range of
\(0\) and \(1\). When analyzing how these benchmark values affect model
performance, it's important to observe the change in Mean Squared Error
(MSE) as the benchmark value increases. The MSE exhibits a gradual
increase as the benchmark value progresses from \(0\) to \(1\).
Assessing this rate of increase is crucial. If the increment is not
substantial, the decision might lean towards retaining low-density
hexagons.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the standardized counts
\end{enumerate}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-stdctsScurve-1.pdf}

\caption{\label{fig-stdctsScurve}Distribution of 2D Euclidean
distances.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\item
  See the distribution of counts
\item
  Take the first quartile
\end{enumerate}

@@

\caption{\label{fig-modelScurve}(a) Full hexagonal grid with bin
centroids, (b) Model constructed in 2D with long edges, and (c) Model
constructed in 2D after removing the long edges.}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the 2D distances
\end{enumerate}

{\centering \includegraphics{paper_files/figure-pdf/fig-distScurve-1.pdf}

\caption{\label{fig-distScurve}Distribution of 2D Euclidean distances.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Find a value which is greater than smallest value
\end{enumerate}

@@

To assess how well our method captures and represents the underlying
structure of the high-D data, Mean Squared Error (MSE) is used. When
computing MSE, total model error (see Section~\ref{sec-summary}) is
divided by the number of observations to make it as a mean value (see
Equation~\ref{eq-equation9}).

@@

@@

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-umapauthor-1.pdf}

@@

\caption{\label{fig-umapauthor}(a) 2D layout from UMAP (n\_neighbors =
30, min\_dist = 0.3) applied for the PBMC3k dataset. Is this a best
representation of the original data?}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-msepltpbmc-1.pdf}

@@

\caption{\label{fig-msepltpbmc}Goodness of fit statistics from UMAP
applied to training PBMC3k dataset. What is the effective number of bins
in UMAP to create a 2D model? The MSE plot have a steep slope at the
beginning, indicating that a smaller number of bins causes a larger
amount of error. Then, the slope gradually declines or level off,
indicating that a higher number of bins generates a smaller error. Using
the elbow method, when the total number of bins is set to 588, the slope
of the MSE plot experiences a sudden and noticeable change, resembling
an elbow-like shape. This point indicates that adding less bins does not
enough to capture the data structure.}

@@

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-72-1.pdf}

The MNIST dataset consists of grayscale images of handwritten digits
\citep{lecun2010}. \citet{Yingfan2021} used this dataset to demonstrate
how the PaCMAP preserves local structure. We select the 2D embedding of
PaCMAP for the handwritten digit 1. As shown in
Figure~\ref{fig-pacmapauthor}, we observe a nonlinear structure present
in 2D space. To answer the following questions,

\begin{itemize}
\tightlist
\item
  Is the original data structure of the handwritten digit 1 dataset
  nonlinear?
\item
  Where does PaCMAP preserve the original data structure?
\item
  Where does PaCMAP show quirks?
\end{itemize}

we visualize the model constructed with PaCMAP overlaid on data in
high-D space (model-in-data space).

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-pacmapauthor-1.pdf}

@@

\caption{\label{fig-pacmapauthor}(a) 2D layout from PaCMAP
(n\_components=2, n\_neighbors=10, init=random, MN\_ratio=0.9,
FP\_ratio=2.0) applied for the digit 1 of the MNIST dataset. Is this a
best representation of the original data?}

@@
